Impute missing values of training and validation data using multiple imputation

### 2022-12-19 Code/11_TestModel.Rmd
Validaiton added

---
title: "Create histogram with risk scores"
author: "Carolien C.H.M. Maas"
date: "14-11-2022"
output: html_document
---

@@

# Test if model is correct
### Check calibration
Intercept of recurrence should be the same as when setting covariates to 'zero' values minus the center
```{r, eval=TRUE}
load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')

int.Rec
20*f.mi.BS.Rec.5$coef["Age.SN"]+
  (log(0.1)-c.Breslow)*f.mi.BS.Rec.5$coef["Breslow"]+
  log(0.01)*f.mi.BS.Rec.5$coef["Rdamcrit"]-
  f.mi.BS.Rec.5$center
```

### Check mean event rates of risk distribution
#### Recurrence
```{r, eval=TRUE}
sum(h.Rec$density*fun.event(lp=lp.points.Rec.5, h0=h0.Rec))
# equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
SF<-survival::survfit(S.Rec.5.training~1,data=training.data.to.be.imputed)
100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
# average probability of recurrence
mean(fun.event(lp=lp.score.Rec, h0=h0.Rec))
```

#### MSM
```{r, eval=TRUE}
sum(h.Rec$density*fun.event(lp=lp.points.MSM.5, h0=h0.MSM))
# equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
SF<-survfit(S.MSM.5.training~1,data=training.data.to.be.imputed)
100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
# average probability of MSM
mean(fun.event(lp=lp.MSM, h0=h0.MSM))
```

>>>>

---
title: "Create histogram with risk scores"
author: "Carolien C.H.M. Maas"
date: "December 2022"
output: html_document
---

@@

# Test if model is correct
### Check calibration
Intercept of recurrence should be the same as when setting covariates to "zero" values minus the center
```{r, eval=TRUE}
load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")
options(datadist="dd.training")

# int.Rec only for last imputed data set
int.Rec
lp.BS <- f.mi.BS.Rec.5$linear
f.mi.BS.MSM.5 <- rms::cph(S.MSM.5.training~lp.BS,y=TRUE,x=TRUE)
20*f.mi.BS.Rec.5$coef["Age.SN"]+
  (log(0.1)-c.Breslow)*f.mi.BS.Rec.5$coef["Breslow"]+
  log(0.01)*f.mi.BS.Rec.5$coef["Rdamcrit"]-
  f.mi.BS.Rec.5$center
```

### Check mean event rates of risk distribution
#### Recurrence
```{r, eval=TRUE}
# h0 for only last imputed data set
f.Rec.basehaz.5 <- survival::basehaz(f.mi.BS.Rec.5,TRUE)
h0.Rec.5 <- f.Rec.basehaz.5$hazard[f.Rec.basehaz.5$time==max(f.Rec.basehaz.5$time[f.Rec.basehaz.5$time<=horizon])]

# MSM.cal.fact for only last imputed data set
lp.5<-predict(f.mi.BS.Rec.5,
              newdata=mice::complete(imputed.training.data,5), 
              type="lp")
f.mi.5 <- rms::cph(S.MSM.5.training~lp.5,y=TRUE,x=TRUE) 
MSM.cal.fact.5 <- stats::coef(f.mi.5)

sum(h.Rec$density*fun.event(lp=lp.points.Rec.5, h0=h0.Rec.5))
# similar to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
SF<-survival::survfit(S.Rec.5.training~1,data=training.data.to.be.imputed)
100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
# average probability of recurrence
mean(fun.event(lp=lp.Rec, h0=h0.Rec.5))
```

#### MSM
```{r, eval=TRUE}
lp.BS <- f.mi.BS.Rec.5$linear
f.mi.BS.MSM.5 <- rms::cph(S.MSM.5.training~lp.BS,y=TRUE,x=TRUE)
f.MSM.basehaz.5 <- survival::basehaz(f.mi.BS.MSM.5,TRUE)
h0.MSM.5 <- f.MSM.basehaz.5$hazard[f.MSM.basehaz.5$time==max(f.MSM.basehaz.5$time[f.MSM.basehaz.5$time<=horizon])]

sum(h.Rec$density*fun.event(lp=lp.points.MSM.5, h0=h0.MSM.5))
# equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
SF<-survfit(S.MSM.5.training~1,data=training.data.to.be.imputed)
100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
# average probability of MSM
mean(fun.event(lp=lp.MSM, h0=h0.MSM.5))
```

### 2022-12-19 Code/12_Validation.Rmd
Validaiton added

# load libraries
library(mice)
library(survival)         # needed to use
# remotes::install_github("CHMMaas/PredictionTools")
library(PredictionTools)  # to plot calibration
# import helper functions
source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
```

# Validation
# Compute predictions for each multiple imputation and make calibration plots with metrics
```{r, eval=TRUE}
# laod data
load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData')
options(datadist='dd.training')
options(digits=8)
load('Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData')
# load model
load('Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata')
p.Rec.val <- matrix(NA, nrow(validation.data.to.be.imputed), m)
p.MSM.val <- matrix(NA, nrow(validation.data.to.be.imputed), m)
for (i in 1:m){
  # Linear predictor
  lp.val.i <- predict(f.mi.BS.Rec.5,
                      newdata=mice::complete(imputed.validation.data, i),
                      type="lp")

  # Recurrence
  p.Rec.val[, i] <- 1-exp(-h0.Rec*exp(lp.val.i))

  # MSM
  p.MSM.val[, i] <- 1-exp(-h0.MSM*exp(MSM.cal.fact*lp.val.i))
}

# calibration curve for recurrence, TODO: based on dd.training?
png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/val.Rec.png"))
PredictionTools::val.surv.mi(p.Rec.val,S.Rec.5.validation,g=5,main="Calibration plot of recurrence for all patients",
                             time=NULL,lim=c(0,1),dist=TRUE)
grDevices::dev.off()

# calibration curve for MSM
png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/val.MSM.png"))
PredictionTools::val.surv.mi(p.MSM.val,S.MSM.5.validation,g=5,main="Calibration plot of MSM for all patients",
                             time=NULL,lim=c(0,1),dist=TRUE)
grDevices::dev.off()

```

>>>>

# load libraries
library(mice)
library(survival)
library(PredictionTools)
source("Z:/Project Melanoom/PaperMelanoma/Code/PaperMelanomaFunctions.R")
```

# Training data
```{r, eval=TRUE}
# load model
load("Z:/Project Melanoom/PaperMelanoma/Results/h0.Rec.MSM.Rdata")

# load data
load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData")

# predictions on training data
p.Rec.5.training <- matrix(NA, nrow(training.data.to.be.imputed), m)
p.MSM.5.training <- matrix(NA, nrow(training.data.to.be.imputed), m)
for (i in 1:m){
  lp.train.i <- predict(f.mi.BS.Rec.5,
                      newdata=mice::complete(imputed.training.data, i),
                      type="lp")
  p.Rec.5.training[, i] <- fun.event(h0=h0.Rec, lp=lp.train.i)
  p.MSM.5.training[, i] <- fun.event(h0=h0.MSM, lp=MSM.cal.fact*lp.train.i)
}
```

# Validation data
```{r, eval=TRUE}
load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData")
p.Rec.5.validation <- matrix(NA, nrow(validation.data.to.be.imputed), m)
p.MSM.5.validation <- matrix(NA, nrow(validation.data.to.be.imputed), m)
for (i in 1:m){
  lp.val.i <- predict(f.mi.BS.Rec.5,
                      newdata=mice::complete(imputed.validation.data, i),
                      type="lp")
  p.Rec.5.validation[, i] <- 1-exp(-h0.Rec*exp(lp.val.i))
  p.MSM.5.validation[, i] <- 1-exp(-h0.MSM*exp(MSM.cal.fact*lp.val.i))
}
```

```{r, eval=TRUE}
# calibration curve for recurrence
# TODO: validation based on dd.training or dd.validation?
for (data.type in c("training", "validation")){
  for (outcome in c("Rec.5", "MSM.5")){
    options(datadist=paste0("dd.", data.type))

    S <- eval(parse(text=paste0("S.", outcome, ".", data.type)))
    p <- eval(parse(text=paste0("p.", outcome, ".", data.type)))
    data <- eval(parse(text=paste0(data.type, ".data.to.be.imputed")))

    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".All.png"))
    par(mar=rep(0, 4))
    PredictionTools::val.surv.mi(p, S,
                                 g=5,main="All patients", time=NULL,lim=c(0,1),dist=TRUE)
    grDevices::dev.off()

    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".Positive.png"))
    par(mar=rep(0, 4))
    PredictionTools::val.surv.mi(p[data$SNstatus=="Positive",],
                                 S[data$SNstatus=="Positive",],
                                 g=5,main="Patients with positive SN", time=NULL,lim=c(0,1),dist=TRUE)
    grDevices::dev.off()

    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".Negative.png"))
    par(mar=rep(0, 4))
    PredictionTools::val.surv.mi(p[data$SNstatus=="Negative",],
                                 S[data$SNstatus=="Negative",],
                                 g=5,main="Patients with negative SN", time=NULL,lim=c(0,1),dist=TRUE)
    grDevices::dev.off()

    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".Neg.Pos.png"), 
        width=32, height=16, units="cm", res=600)
    par(mar=rep(0, 4))
    layout(matrix(1:2, ncol=2, byrow=TRUE))
    for (select.value in c("Positive", "Negative")){
      plot(NA, xlim=0:1, ylim=0:1, xaxt="n", yaxt="n", bty="n")
      img <- png::readPNG(paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".", select.value, ".png"))
      rasterImage(img, 0, 0, 1, 1)
    }
    grDevices::dev.off()

    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/", data.type, ".", outcome, ".png"), width=32, height=45, units="cm", res=600)
    par(mar=rep(0, 4))
    layout(matrix(1:2, ncol=1, nrow=2, byrow=TRUE), heights=c(2, 1), widths=c(2, 2))
    for (select.value in c("All", "Neg.Pos")){
      plot(NA, xlim=0:1, ylim=0:1, xaxt="n", yaxt="n", bty="n")
      img <- png::readPNG(paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/", data.type, ".", outcome, ".", select.value, ".png"))
      rasterImage(img, 0, 0, 1, 1)
    }
    grDevices::dev.off()
  }
}
```

### 2022-12-19 Code/1_DescriptivesTable.Rmd
Validaiton added

```{r setup, include=FALSE}
# clear all
rm(list = ls(all.names = TRUE))
if(!is.null(dev.list())) dev.off()
cat('\014')

@@

load('Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata')
load('Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata')
descr.table <- arsenal::tableby(Center ~ SNstatus
                                +Sex
                                +Age.SN
                                +arsenal::includeNA(Ulceration)
                                +arsenal::includeNA(Loc_CAT)
                                +arsenal::includeNA(Histology)
                                +Breslow
                                +multiple.fields
                                +Tot_SNs_neg
                                +Tot_SNs_pos
                                +Rdamcrit
                                +arsenal::includeNA(Dewar)
                                +arsenal::includeNA(Mitosis)
                                +Recurrence
                                +dead
                                +MSM
                                +logRec.time
                                +logFU.time, data=rbind(training.data.to.be.imputed, validation.data.to.be.imputed), 
                                total.pos="before",
                                digits = 1,
                                test = F,
                                total = T,
                                numeric.test = "kwt", cat.test = "chisq",
                                numeric.stats = c("meansd", "medianq1q3", "range", "Nmiss2"),
                                cat.stats = c("countpct"),
                                stats.labels = list(
                                  meansd = "Mean (SD)",
                                  medianq1q3 = "Median (Q1, Q3)",
                                  range = "Min - Max",
                                  Nmiss2 = "Missing"
                                ))
my_labels <- list(
  SNstatus = "SN status",
  Age.SN = "Age",
  Loc_CAT = "Location",
  multiple.fields = "Multiple fields",
  Tot_SNs_neg = "Total number of negative SN",
  Tot_SNs_pos = "Total number of positive SN",
  Rdamcrit = "SN tumour burden",
  Dewar = "Location metastasis in lymph",
  dead = "Deaths",
  MSM = "Melanoma specific mortality"
)
```

>>>>

```{r setup, include=FALSE}
# clear all
rm(list = ls(all.names = TRUE))
if(!is.null(dev.list())) dev.off()
cat("\014")

@@

load("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata")
load("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.Rdata")
descr.table <- arsenal::tableby(Center ~ SNstatus
                                +Sex
                                +Age.SN
                                +arsenal::includeNA(Ulceration)
                                +arsenal::includeNA(Loc_CAT)
                                +arsenal::includeNA(Histology)
                                +Breslow
                                +multiple.fields
                                +Tot_SNs_neg
                                +Tot_SNs_pos
                                +Rdamcrit
                                +arsenal::includeNA(Dewar)
                                +arsenal::includeNA(Mitosis)
                                +Recurrence
                                +dead
                                +MSM
                                +logRec.time
                                +logFU.time, data=rbind(training.data.to.be.imputed, validation.data.to.be.imputed[, -c(18, 19)]), # training.data.to.be.imputed,
                                total.pos="before",
                                digits = 1,
                                test = F,
                                total = T,
                                numeric.test = "kwt", cat.test = "chisq",
                                numeric.stats = c("meansd", "medianq1q3", "range", "Nmiss2"),
                                cat.stats = c("countpct"),
                                stats.labels = list(
                                  meansd = "Mean (SD)",
                                  medianq1q3 = "Median (Q1, Q3)",
                                  range = "Min - Max",
                                  Nmiss2 = "Missing"
                                ))
my_labels <- list(
  SNstatus = "SN status",
  Age.SN = "Age",
  Loc_CAT = "Location",
  multiple.fields = "Multiple fields",
  Tot_SNs_neg = "Total number of negative SN",
  Tot_SNs_pos = "Total number of positive SN",
  Rdamcrit = "SN tumour burden",
  Dewar = "Location metastasis in lymph",
  dead = "Deaths",
  MSM = "Melanoma specific mortality"
)
```

### 2022-12-19 Code/2_ImputeData.Rmd
Validaiton added

---
title: "Impute missing values of training and validation data using multiple imputation"
author: "Carolien C.H.M. Maas"
date: "14-11-2022"
output: html_document
---

@@

Function to perform multiple imputation for training or validation data
```{r, eval=TRUE}
impute.missing.values <- function(type=NULL, m=1, new.imputation=FALSE){
  # load data
  load(paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/', type, '.data.Rdata'))

  if (new.imputation){
    # impute data
    data.to.be.imputed <- eval(parse(text=paste0(type, '.data.to.be.imputed')))
    imputed.data <- mice::mice(data.to.be.imputed, m=m, maxit=5, print=FALSE, seed=500)
    assign(paste0('imputed.', type, '.data'), imputed.data)
    save(list=c(paste0('imputed.', type, '.data')),
         file=paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/imputed.', type, '.data.Rdata'), compress=TRUE)
  }
  load(paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/imputed.', type, '.data.Rdata'))
  save(list=c(paste0('dd.', type), paste0(type, '.data.to.be.imputed'), 'm', paste0('imputed.', type, '.data')),
       file=paste0('Z:/Project Melanoom/PaperMelanoma/Data/', type, '/', type, '.data.Rdata'), compress=TRUE)
}
```

Perform multiple imputation
```{r, eval=TRUE}
# erase memory
m <- 5
new.imputation <- FALSE
impute.missing.values(type='training', m=m, new.imputation=new.imputation)
impute.missing.values(type='validation', m=m, new.imputation=new.imputation)

>>>>

---
title: "Impute missing values of training and validation data using multiple imputation"
author: "Carolien C.H.M. Maas"
date: "December 2022"
output: html_document
---

@@

Function to perform multiple imputation for training or validation data
```{r, eval=TRUE}
impute.missing.values <- function(type=NULL, m=1, new.imputation=FALSE){
  # load data
  load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/", type, ".data.Rdata"))

  if (new.imputation){
    # impute data
    data.to.be.imputed <- eval(parse(text=paste0(type, ".data.to.be.imputed")))
    imputed.data <- mice::mice(data.to.be.imputed, m=m, maxit=5, print=FALSE, seed=500)
    # print(imputed.data$method)
    # print(imputed.data$loggedEvents)
    assign(paste0("imputed.", type, ".data"), imputed.data)
    save(list=c(paste0("imputed.", type, ".data")),
         file=paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/imputed.", type, ".data.Rdata"), compress=TRUE)
  }
  load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/imputed.", type, ".data.Rdata"))
  save(list=c(paste0("dd.", type), paste0(type, ".data.to.be.imputed"), "m", paste0("imputed.", type, ".data")),
       file=paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/", type, ".data.Rdata"), compress=TRUE)
}
```

Perform multiple imputation
```{r, eval=TRUE}
# erase memory
m <- 5
new.imputation <- TRUE
# Impute: Ulceration, Loc_CAT, Histology, Breslow, Tot_SNs_neg, Tot_SNs_pos, Rdamcrit, Dewar, Mitosis, logRec.time
impute.missing.values(type="training", m=m, new.imputation=new.imputation)
# Impute: Ulceration, Histology, Rdamcrit, Dewar, logRec.time
impute.missing.values(type="validation", m=m, new.imputation=new.imputation)
